Artificial ecosystem selection reveals relationships between microbiome composition and ecosystem function

### 2023-03-03 manuscript/manuscript.Rmd
incorporate final edits

{title max characters 150; current = 106 w/spaces}  
^1^ Institute of Ecology and Evolution, University of Oregon, Eugene, OR, USA  
\* Correspondence: Andrew H. Morris, amorris3\@uoregon.edu  
Competing Interests: We declare we have no competing interests.

@@

```{r initialize-packages}
library(tidyverse)
library(broom)
library(broman)
library(morris)
# library(phyloseq)
library(ggpubr)
library(knitr)
library(corncob)
# library(ggthemes)
library(captioner)
library(knitr)
library(ggtext)
library(patchwork)
figures <- captioner()
tables <- captioner(prefix = "Table")

@@

```{r load-data}
heritability <- read_tsv(paste0(der_dir, 'heritability.tsv'))
fluxes <- read_tsv(paste0(der_dir, 'fluxes.tsv'))

@@

#load(file = paste0(der_dir, 'plots.RData'))
load(file = paste0(der_dir, 'models.RData'))

# # CH4 models
# load(file = paste0(der_dir, 'response_model.RData'))
# load(file = paste0(der_dir, 'response_plot.RData'))
# load(file = paste0(der_dir, 'herit_model.RData'))
# load(file = paste0(der_dir, 'herit_plot.RData'))
# 
# # 16S models
# load(file = paste0(der_dir, 'richness_models.RData'))
# load(file = paste0(der_dir, 'phyloseq.RData'))
# load(file = paste0(der_dir, 'beta_model.RData'))
# load(file = paste0(der_dir, 'beta_fig.RData'))
# load(file = paste0(der_dir, 'da_corncob.RData'))
# load(file = paste0(der_dir, 'da_fig.RData'))
# load(file = paste0(der_dir, 'dbrda.RData'))

@@

{max 250 words; current = 249}  
Microbiomes mediate important ecosystem functions, yet it has proven difficult
to determine the relationship between microbiome composition and the rate of
ecosystem functions. This challenge remains because it is difficult to
manipulate microbiome composition directly, we often cannot know _a priori_ which
microbiome members influence the rate of an ecosystem function, and
microbiomes can covary strongly with other drivers of ecosystem function, such
as the environment. We propose that artificial ecosystem selection can overcome
these challenges. Artificial ecosystem selection involves creating replicate
ecosystems, selecting replicates with a desired ecosystem-level trait, and
using the microbiomes from the selected ecosystems to inoculate a new set of
ecosystems. We performed artificial ecosystem selection on replicate soil
microbiomes to select for increased soil methane oxidation rate. We observed a
strong response to selection with a `r myround(per_slope, 1)`\% increase in
methane oxidation rate per passage. We
estimated that `r myround(tidy(rh_model)[[2, 2]]*100, 1)`\% of the variation in
methane oxidation
rate in these soils can be attributed to microbiome variation. We also found
that selection did not enrich for known methane oxidizers; instead, `r length(da_fams)` families not
known to oxidize methane, including *Fimbriimonadaceae*, *Cytophagaceae*, and
*Diplorickettsiaceae*, were enriched by selection. This result is in contrast to the typical assumption that
the rate of an ecosystem function is limited by the final step in the associated
microbial pathway. Our study demonstrates that variation in microbiome
composition can contribute to variation in the rate of ecosystem function and
that manipulating microbiome composition could be a viable strategy for managing
ecosystem functions.

{max text 5000 words; current text = 4549}

@@

(`r figures("concept", display = "cite")`)

There have been two general categories of approaches that investigators have
used to estimate the degree to which an ecosystem function varies with
microbiome composition: comparative and manipulative. Comparative
studies sample natural variation in an ecosystem function across different
habitats and simultaneously measure variation in community composition.
Investigators can then correlate ecosystem function with aspects of community
composition while attempting to control for environmental variation. These
approaches have
documented important relationships between microbiomes and ecosystem
functions. For example, a meta-analysis of these studies observed a small but
significant contribution of the microbiome to variation in ecosystem
function after controlling for environmental variation [@graham2016]. In
addition, studies focusing on the correlation between the rate of an ecosystem
function and the abundance of an associated marker gene (i.e., a gene that codes
for a protein assumed to be involved in the ecosystem function) sometimes observe a
significant correlation, though this relationship is rare and contingent upon
both the function and the ecosystem sampled [@rocca2015]. However, comparative
studies come with some unique challenges and limitations. One issue is that
microbiome attributes tend to covary with the abiotic conditions within an
environment, and whether and how the investigator controls for these variables
can affect the conclusions of such a comparative study. In addition, it is
difficult to know _a priori_ which environmental variables or community attributes
to measure. Finally, while these approaches can establish a potential magnitude
and direction for these relationships, it is often difficult to identify the
taxa or genes that explain the connection between composition and function.

The other broad category of approaches used to address this question are
manipulative approaches. Manipulative experiments try to alter microbial
community composition and observe the effect on function. For example,
reciprocal transplant and common garden experiments have shown that microbiomes
originating from different ecosystems inoculated into the same substrate or
introduced into a common environment display distinct functional rates 
[@cavigelli2000; @balser2005; @waldrop2006; @strickland2009]. In addition,
manipulating diversity by filtering communities by cell size or through dilution
has been shown to alter the rate of ecosystem functions 
[@wagg2014; @wagg2019; @griffiths2001]. However, manipulating the microbiome
directly is challenging. Each of the manipulative approaches applied confound
community composition with other factors. For example, reciprocal transplant and
common garden experiments can confound community composition with the abiotic
conditions introduced with the inoculum, while manipulating composition through
dilution may confound composition with biomass [@reed2007].

In this study, we sought to build on the observations of comparative and
manipulative studies by applying a different approach to the question of whether
microbiome variation contributes to variation in the rate of an ecosystem
function. We used artificial ecosystem selection to select for microbiomes that
performed a greater rate of ecosystem function
[@swenson2000; @blouin2015; @panke-buisse2015]. We then tested whether variation
in the microbiome contributed to variation in the rate of ecosystem function and
identified microbiome attributes that might explain this relationship. There are
several potential advantages to this approach for documenting
microbiome-function relationships and for investigating the mechanisms
underlying those relationships. By passaging microbiomes over multiple rounds
using a common environment, we can weaken the covariance between microbes and
the environment by repeatedly diluting abiotic effects. In addition, by
inoculating our ecosystems with equally-sized subsamples, we eliminate the need
to generate microbiome variation through methods that are confounded with
biomass or cell size. Lastly, by comparing our artificially selected community
to a control community resulting from random selection, we can both control for
changes in the environment over time and identify genes or taxa that are
associated with the ecosystem function under selection.

We applied artificial ecosystem selection to soil microbiomes by selecting on
soil methane oxidation rate. We chose this function because methane is a
globally important greenhouse gas and methane oxidation by soil bacteria is the
primary biological sink for atmospheric methane [@kirschke2013]. In addition,
there is evidence that soil methane oxidation rate may vary with microbiome
composition based on comparative studies in a variety of arctic and tropical
ecosystems [@meyer2020; @meyer2020a; @freitag2009; @freitag2010] as well as
studies that manipulate methanotroph richness [@schnyder2018]. Finally,
methanotrophy is one of the most deeply conserved microbial physiologies and is
represented in a narrow range of taxa, which suggests that the taxonomic
composition of the microbiome is more likely to be associated with the rate of
methane oxidation than other broader or more shallowly conserved functions
[@schimel1998; @martiny2013].

In this study, we used artificial ecosystem selection on methane oxidation rate
to address the following questions: Does variation in the relative abundance of
microbial taxa contribute to variation in soil methane oxidation rate in our
system? Which attributes of the microbiome are associated with variation in
methane oxidation rate, and do these attributes match our assumptions about the
factors that regulate methane oxidation rate in nature?

@@

We performed an artificial ecosystem selection experiment 
[*sensu* @swenson2000] by passaging replicate soil microbiomes. The trait we
selected on was methane (CH~4~)-oxidation rate. This experiment had two selection lines
with twelve jars each for a total of twenty-four jars per passage. One line was
a positive selection line where the two or three jars with the highest CH~4~
oxidation rate were homogenized to inoculate the next set of positive jars. The
other line was a neutral selection line where an equal number of jars as the
positive line were chosen at random to inoculate the next set of neutral jars.
The number of jars chosen was based on the distribution of fluxes among the
positive jars, i.e., we chose the top three jars unless only two jars had
considerably greater CH~4~ oxidation rates based on visual inspection of
histograms. The experiment was carried out over five passages until a
significant divergence in functional rates was observed between the two
selection treatments based on a difference of slopes.

The initial soil microbiome was sampled from the top 10 cm of an upland mineral
soil under a deciduous forest ecosystem near the University of Oregon campus in
Eugene, OR, USA. Incubations were performed in 500 mL mason jars with rubber
septa installed in the lids. Each jar was sterilized with 70\% ethanol to which
was added 45 g of autoclaved potting mix, 5 g of living soil, and 3.5 mL of
sterile deionized water to bring them to 60\% of field capacity. The potting
mix was Lane Potting Mix from Lane Forest Products (Eugene, OR, USA). Each jar
was then capped and injected with 4.3 mL of 99\% CH~4~ to bring
the headspace concentration to approximately 1000 ppm CH~4~. To create the two
treatment groups, twenty-four jars were created in an identical manner and then
randomly assigned to either the positive or neutral selection line. Jars were
flushed and respiked with CH~4~ twice per week to maintain aerobic conditions
and elevated CH~4~ concentrations and were incubated at ambient temperature for
approximately four weeks. Methane oxidation rates were determined at the end of
the incubation period. For the positive treatment, the three jars with the
greatest CH~4~ oxidation rates were chosen to inoculate the next generation.
For the neutral treatment, three jars were randomly selected to inoculate the
next generation. For each treatment, the selected jars were homogenized and 5 g
of the homogenized soil was used to inoculate the next set of jars, which
represents a 10% subsampling. The next set of jars were created in an identical
manner to the first generation with fresh autoclaved potting mix and the same
moisture and CH~4~ content.

@@

Methane oxidation rates were determined after flushing and spiking jars to
approximately 1000 ppm CH~4~. Headspace samples of 1 mL were collected from
each jar immediately after spiking and then at time points 3, 6, 24, and 48
hours for a 5-point curve. Samples were immediately injected into a SRI model
8610C gas chromatograph equipped with a flame ionization detector (SRI
Instruments, Torrance, CA, USA) to determine the headspace CH~4~ concentration.
We applied a first-order exponential decay function to determine the rate
constant (k, units = d^−1^; i.e., dCH~4~/dt = k[CH~4~]) of the exponential
decrease in methane. Oxidation rates are presented as the additive inverse of
$k$ (i.e., $-k$) so that a more positive value represents a greater
oxidation rate. The jars selected in passage two for the positive treatment had
the lowest CH~4~ oxidation rate of the twelve jars due to a calculation error
in the rate constant. Despite this error, we still observed a strong
response to selection demonstrating the power of this technique. In addition,
this likely reintroduced diversity lost during the selection process,
potentially increasing the variation available to be selected upon, as
is recommended for microbiome selection experiments [@sanchez2021].

@@

Statistical analyses were performed in 'R' [@rcoreteam2018]. To determine
whether there was a significant change in CH~4~ oxidation rate as a response to
selection, we tested a difference in slopes between the positive and neutral
selection lines. Residuals did not meet the assumptions of constant variance
and normal distribution. Therefore, CH~4~ oxidation rates were log~10~
transformed prior to analysis. First, we tested if there was a difference of
slopes between the positive and neutral treatments based on the interaction
between passage and treatment. To test the interaction, we fit two nested models
with and without the interaction term and compared them using an F-test with the
'anova' function. We then present the slopes for each treatment, which
represented the change in CH~4~ oxidation rate per passage as a response to
selection.

We estimated the proportion of variation in methane oxidation rate due to
variation in the microbiome as the regression of divergence between the positive
and neutral treatment on the cumulative selection differential [@falconer1996].
Divergence was calculated as the mean methane oxidation rate of the positive
treatment minus the mean methane oxidation rate of the neutral treatment in each
passage. The selection differential was calculated as the difference between
the mean of the three selected jars and the mean of all twelve jars in a passage.
Cumulative selection differential was calculated as the sum of the selection
differential from all preceding selection events. We then regressed cumulative
divergence on cumulative selection differential using the 'lm' function.

@@

We observed a response to artificial selection on whole-ecosystem soil CH~4~ oxidation
rate (`r figures("response", display = "cite")`; difference of slopes:
F~`r compare_response_mods[2,4]`,`r compare_response_mods[2,2]`~ = 
`r myround(compare_response_mods[2,5], 2)`,
p = `r myround(compare_response_mods[2,6], 2)`). At the start of the experiment,
the Positive treatment had a mean CH~4~ oxidation rate that was
`r percent_less(initial = tidy(response_model)[[1, 2]], final = tidy(response_model)[[3, 2]] + tidy(response_model)[[1, 2]])`\%
lower than the Neutral treatment
`r print_lm(response_model, 3, 'difference of y-intercepts')`. 
There was no change in CH~4~ oxidation rate of the Neutral treatment over the
five passages `r print_lm(response_model, 2)`. By contrast, the Positive
treatment had a `r myround(per_slope, 1)`\% increase in CH~4~ oxidation rate per
passage `r print_lm(response_model, 4)`.

@@

Median ASV richness decreased from `r myround(rare_median[1, 2], 1)` (`r myround(rare_median[1, 3], 1)`) in Passage 2 to `r myround(rare_median[2, 2], 1)` (`r myround(rare_median[2, 3], 1)`) in Passage 5 (Kruskal-Wallace test: $\chi^2$ = 
`r myround(kt[1,1], 1)`, df = `r kt[1,3]`, p < 0.001; pairwise Wilcoxon test: p < 0.001). However, there was no
difference in richness between the Positive and the Neutral treatment in
Passage 2 or 5 (pairwise Wilcoxon test: p = `r myround(wt[1, 3], 1)`).  In addition, there was no
correlation between richness and CH~4~ oxidation rate across the two treatments in Passage 5 (Spearman's rho = `r myround(s5[1,1], 1)`, p = `r myround(s5[1,3], 1)`).

@@

Bray-Curtis dissimilarity of the soil microbiome varied strongly by passage and
weakly by treatment with an interaction between passage and treatment (`r figures("beta", display = "cite")`). 
Passage explained `r myround(100*tidy_permanova[1, 4])`\% of the
variation in Bray-Curtis dissimilarity (F~`r tidy_permanova[1, 2]`,`r tidy_permanova[4, 2]`~ = `r myround(tidy_permanova[1, 5], 1)`, p = `r myround(tidy_permanova[1, 6], 3)`), treatment explained 
`r myround(100*tidy_permanova[2, 4])`\%
of the variation (F~`r tidy_permanova[2, 2]`,`r tidy_permanova[4, 2]`~ = `r myround(tidy_permanova[2, 5], 1)`, p = `r myround(tidy_permanova[2, 6], 3)`), and the interaction
between treatment and passage explained 
`r myround(100*tidy_permanova[3, 4])`\% of the variation (F~`r tidy_permanova[3, 2]`,`r tidy_permanova[4, 2]`~ = `r myround(tidy_permanova[3, 5], 1)`, p = `r myround(tidy_permanova[3, 6], 3)`). There was 
no difference in dispersion between treatments or passages (F~3,44~ = 0.91, p = 0.45). 
Finally, CH~4~ oxidation rate was correlated with Bray-Curtis dissimilarity across both treatments in Passage 5 and explained `r myround(dbrda_percent_explained[1], 1)`\% of the variation in Bray-Curtis dissimilarity (dbRDA: F~`r dbrda_aov[1, 1]`,`r dbrda_aov[2, 1]`~ = `r myround(dbrda_aov[1, 3], 2)`, p = `r myround(dbrda_aov[1, 4], 3)`)

@@

To identify taxa that responded
to selection on soil CH~4~ oxidation rate, we tested the differential relative
abundance of families in the Positive jars relative to the Neutral jars within
Passage 5 using three methods and then plotted the taxa identified by all three methods. We identified `r length(da_fams)`
families that were enriched or depleted in the Positive treatment relative to
the Neutral treatment (`r figures("da", display = "cite")`).

Overall, none of the families enriched in the Positive selection treatment
contain known methanotrophs. Several taxa identified had a 
higher taxonomic designation that contains methanotrophs, for example, the
Gammaproteobacteria class had a large effect size.
The Gammaproteobacteria include the type I and type X methanotrophs in the families *Methylococcaceae* and *Methylothermaceae* [@stein2012].
However, the Gammaproteobacteria is among the most diverse 
groups in the Prokaryotes, so this is not strong evidence for a selection response
by methanotrophs [@garrity2005].
In addition, the *Puniceicoccaceae* is a member of the phylum Verrucomicrobia. The Verrucomicrobia is a diverse group that contain
known methanotrophs as well as ammonia-oxidizing bacteria [@freitag2003]. Other than these two groups, none of the other taxa enriched in the Positive treatment are known to be related to methanotrophs. Two groups in the Armatimonadales were enriched in the Positive treatment including the family *Fimbriimonadaceae* and an unclassified ASV from the order Armatimonadales [@im2012].
*Cytophagaceae* was also enriched in
the Positive treatment and contains a number of mainly aerobic heterotrophs
that can digest a variety of macromolecules [@mcbride2014]. 
The remaining families include the uncultured family 0319-6G20, *Diplorickettsiaceae*, *Rhodospirillaceae*, and an unclassified Kapabacteriales.

We did not identify any methanotrophic families in the overall differential abundance analysis. 
However, we wanted to look more closely at the known methanotrophs in our dataset to be sure that they did not have an effect.
To do this, we subset the ASVs in our dataset that were in families that contained methanotrophs.
Only two families were represented: *Methylacidiphilaceae* and *Beijerinckiaceae*. 
Aggregating reads at the family level, neither family was differentially abundant between the two treatments.
However, aggregated at the genus level, a group of unclassified genera in the *Beijerinckiaceae* were depleted in the Positive treatment and the genus
*Rhodoblastus*, a member of the *Beijerinckiaceae*, was enriched in the
Positive treatment.  While many *Beijerinckiaceae* are methanotrophs, several
taxa in this family have lost the ability to oxidize CH~4~ and it appears
that *Rhodoblastus* species are not able to grow on CH~4~, though they can grow
on methanol [@tamas2014; @dedysh2016]. Based on this analysis,
it appears that no methanotrophs were enriched in the Positive treatment.

>>>>

^1^ Institute of Ecology and Evolution, University of Oregon, Eugene, OR, USA  
\* Correspondence: Andrew H. Morris, amorris3\@uoregon.edu  
Competing Interests: We declare we have no competing interests.

@@

```{r initialize-packages}
library(tidyverse)
library(broom)
library(broman)
library(morris)
library(ggpubr)
library(knitr)
library(corncob)
library(captioner)
library(knitr)
library(ggtext)
library(patchwork)
figures <- captioner()
tables <- captioner(prefix = "Table")

@@

```{r load-data}
#heritability <- read_tsv(paste0(der_dir, 'heritability.tsv'))
#fluxes <- read_tsv(paste0(der_dir, 'fluxes.tsv'))

@@

load(file = paste0(der_dir, 'models.RData'))

@@

Microbiomes mediate important ecosystem functions, yet it has proven difficult
to determine the relationship between microbiome composition and the rate of
ecosystem functions. This challenge remains because it is difficult to
manipulate microbiome composition directly, we often cannot know _a priori_ which
microbiome members influence the rate of an ecosystem function, and
microbiomes can covary strongly with other drivers of ecosystem function, such
as the environment. We propose that artificial ecosystem selection can overcome
these challenges. Artificial ecosystem selection involves creating replicate
ecosystems, selecting replicates with a desired ecosystem-level trait, and
using the microbiomes from the selected ecosystems to inoculate a new set of
ecosystems. We performed artificial ecosystem selection on replicate soil
microbiomes to select for increased soil CH~4~ oxidation rate. We observed a
strong response to selection with a `r myround(per_slope, 1)`\% increase in
CH~4~ oxidation rate per passage. We
estimated that `r myround(tidy(rh_model)[[2, 2]]*100, 1)`\% of the variation in
CH~4~ oxidation
rate in these soils can be attributed to microbiome variation. We also found
that selection did not enrich for known CH~4~ oxidizers; instead, `r length(da_fams)` families not
known to oxidize CH~4~, including *Fimbriimonadaceae*, *Cytophagaceae*, and
*Diplorickettsiaceae*, were enriched by selection. This result is in contrast to the typical assumption that
the rate of an ecosystem function is limited by the final step in the associated
microbial pathway. Our study demonstrates that variation in microbiome
composition can contribute to variation in the rate of ecosystem function and
that manipulating microbiome composition could be a viable strategy for managing
ecosystem functions.

@@

Altering ecosystem functions via microbiome manipulations requires that the
microbiome contributes to variation in ecosystem function independent of other
drivers of ecosystem variation, such as variation in environmental conditions.
This is because the drivers of variation in ecosystem functions can interact in
complicated ways (`r figures("concept", display = "cite")`). Variation in the
microbiome can contribute directly to variation in ecosystem function, for
example, if a microbial population is replaced by one with a greater enzyme
efficiency. In addition, environmental conditions can contribute indirectly to
ecosystem function via covariance with the microbiome, for example, by providing
conditions that select for microbial groups that in turn alter the rates of
ecosystem functions. In this scenario, identifying the change in microbial
community composition without adequately controlling for the environmental
conditions would incorrectly attribute the change in ecosystem function to the
microbiome when it is ultimately an indirect effect of the environment.
Determining the independent contribution of microbiome variation to ecosystem
function is crucial because if microbiome composition is driven primarily by
environmental conditions, then introducing a desirable taxon through microbiome
manipulation without altering the environment will likely be unsuccessful at
shifting the targeted ecosystem function.

There have been two general categories of approaches that investigators have
used to estimate the degree to which an ecosystem function varies with
microbiome composition: comparative and manipulative. Comparative studies sample
natural variation in an ecosystem function across different habitats and
simultaneously measure variation in community composition.  Investigators can
then correlate ecosystem function with aspects of community composition while
attempting to control for environmental variation. These approaches have
documented important relationships between microbiomes and ecosystem functions.
For example, a meta-analysis of these studies observed a small but significant
contribution of the microbiome to variation in ecosystem function after
controlling for environmental variation [@graham2016]. In addition, studies
focusing on the correlation between the rate of an ecosystem function and the
abundance of an associated marker gene (i.e., a gene that codes for a protein
assumed to be involved in the ecosystem function) sometimes observe a
significant correlation, though this relationship is rare and contingent upon
both the function and the ecosystem sampled [@rocca2015]. However, comparative
studies come with unique challenges and limitations. One issue is that
microbiome attributes tend to covary with the abiotic conditions within an
environment, and it is difficult to control for these abiotic variables in order
to identify the unique contribution of the microbiome to ecosystem function. In
addition, it is difficult to know _a priori_ which environmental variables or
community attributes to measure. Finally, while these approaches can establish a
potential magnitude and direction for these relationships, it is often difficult
to identify the
taxa or genes that explain the connection between composition and function.

The other broad category of approaches used to address this question are
manipulative approaches. Manipulative experiments try to alter microbial
community composition and observe the effect on function. For example,
reciprocal transplant and common garden experiments have shown that microbiomes
originating from different ecosystems inoculated into the same substrate or
introduced into a common environment display distinct functional rates 
[@cavigelli2000; @balser2005; @waldrop2006; @strickland2009]. In addition,
manipulating diversity by filtering communities by cell size or through dilution
has been shown to alter the rate of ecosystem functions 
[@wagg2014; @wagg2019; @griffiths2001]. However, manipulating the microbiome
directly is challenging, and manipulative approaches often confound
community composition with other factors. For example, reciprocal transplant and
common garden experiments can confound community composition with the abiotic
conditions introduced with the inoculum, while manipulating composition through
dilution may confound composition with biomass [@reed2007].

In this study, we sought to build on the observations of comparative and
manipulative studies by applying a different approach to the question of whether
microbiome variation contributes to variation in the rate of an ecosystem
function. We used artificial ecosystem selection to select for microbiomes that
performed a greater rate of ecosystem function [@swenson2000; @blouin2015;
@panke-buisse2015]. We then tested whether variation in the microbiome
contributed to variation in the rate of ecosystem function and identified
microbiome attributes that might explain this relationship. There are several
potential advantages to this approach for documenting the direct contribution of
the microbiome to variation in ecosystem function and for investigating the
mechanisms underlying those relationships. By repeated passaging of microbiomes
in a common environment, we can weaken the covariance between microbes and the
environment by repeatedly diluting the influence of variation in abiotic
conditions. In addition, our approach eliminates the need to generate microbiome
variation through methods that are confounded with biomass or cell size. Lastly,
by comparing our artificially selected community to a control community
resulting from random selection, we can both control for changes in the
environment over time and identify genes or taxa that are associated with the
ecosystem function under selection.

We applied artificial ecosystem selection to soil microbiomes by selecting on
soil methane (CH~4~) oxidation rate. We chose this function because CH~4~ is a
globally important greenhouse gas and CH~4~ oxidation by soil bacteria is the
primary biological sink for atmospheric CH~4~ [@kirschke2013]. In addition,
there is evidence that soil CH~4~ oxidation rate may vary with microbiome
composition based on comparative studies in a variety of arctic and tropical
ecosystems [@meyer2020; @meyer2020a; @freitag2009; @freitag2010] as well as
studies that manipulate methanotroph richness [@schnyder2018]. Finally,
methanotrophy is one of the most deeply conserved microbial physiologies and is
represented in a narrow range of taxa, which suggests that the taxonomic
composition of the microbiome is more likely to be associated with the rate of
CH~4~ oxidation than other broader or more shallowly conserved functions
[@schimel1998; @martiny2013].

In this study, we used artificial ecosystem selection on CH~4~ oxidation rate
to address the following questions: Does variation in the relative abundance of
microbial taxa contribute to variation in soil CH~4~ oxidation rate
independent of the environment in our system? Which attributes of the microbiome
are associated with variation in CH~4~ oxidation rate, and do these attributes
match our assumptions about the factors that regulate CH~4~ oxidation rate in
nature?

@@

We performed an artificial ecosystem selection experiment 
(*sensu* @swenson2000) by passaging replicate soil microbiomes. The trait we
selected on was CH~4~ oxidation rate. This experiment had a selection line and a
control line with twelve jars each for a total of twenty-four jars per passage.
The selection line underwent positive selection where the two or three jars
with the highest CH~4~ oxidation rate were homogenized to inoculate the next set
of jars. The control line underwent random selection where an equal number
of jars as the selection line were chosen at random to inoculate the next set of
jars. The number of jars chosen was based on the distribution of fluxes among
the positive jars, i.e., we chose the top three jars unless only two jars had
considerably greater CH~4~ oxidation rates based on visual inspection of
histograms. The experiment was carried out over five passages until a
significant divergence in functional rates was observed between the two
selection treatments based on a difference of slopes.

The initial soil microbiome was sampled from the top 10 cm of an upland mineral
soil under a deciduous forest ecosystem near the University of Oregon campus in
Eugene, OR, USA. Incubations were performed in 500 mL mason jars with rubber
septa installed in the lids. Each jar was sterilized with 70\% ethanol to which
was added 45 g of autoclaved potting mix, 5 g of living soil, and 3.5 mL of
sterile deionized water to bring them to 60\% of field capacity. The potting
mix was Lane Potting Mix from Lane Forest Products (Eugene, OR, USA). Each jar
was then capped and injected with 4.3 mL of 99\% CH~4~ to bring
the headspace concentration to approximately 1000 ppm CH~4~. To create the two
treatment groups, twenty-four jars were created in an identical manner and then
randomly assigned to either the selection or control line. Jars were
flushed and respiked with CH~4~ twice per week to maintain aerobic conditions
and elevated CH~4~ concentrations and were incubated at ambient temperature for
approximately four weeks. Methane oxidation rates were determined at the end of
the incubation period and selection was performed. For each treatment, the
selected jars were homogenized and 5 g of the homogenized soil was used to
inoculate the next set of jars, which represents a 10% subsampling. The next set
of jars were created in an identical manner to the first generation with fresh
autoclaved potting mix and the same moisture and CH~4~ content.

@@

Methane oxidation rates were determined after flushing and spiking jars to
approximately 1000 ppm CH~4~. Headspace samples of 1 mL were collected from
each jar immediately after spiking and then at time points 3, 6, 24, and 48
hours for a 5-point curve. Samples were immediately injected into a SRI model
8610C gas chromatograph equipped with a flame ionization detector (SRI
Instruments, Torrance, CA, USA) to determine the headspace CH~4~ concentration.
We applied a first-order exponential decay function to determine the rate
constant (k, units = d^−1^; i.e., dCH~4~/dt = k[CH~4~]) of the exponential
decrease in CH~4~ Oxidation rates are presented as the additive inverse of
$k$ (i.e., $-k$) so that a more positive value represents a greater
oxidation rate. The jars selected in passage two for the positive selection treatment had
the lowest CH~4~ oxidation rate of the twelve jars due to a calculation error
in the rate constant. Despite this error, we still observed a strong
response to selection demonstrating the power of this technique. In addition,
this likely reintroduced diversity lost during the selection process,
potentially increasing the variation available to be selected upon, as
is recommended for microbiome selection experiments [@sanchez2021].

@@

Statistical analyses were performed in 'R' [@rcoreteam2018]. To determine
whether there was a significant change in CH~4~ oxidation rate as a response to
selection, we tested a difference in slopes between the selection and control
lines. Residuals did not meet the assumptions of constant variance
and normal distribution. Therefore, CH~4~ oxidation rates were log~10~
transformed prior to analysis. First, we tested if there was a difference of
slopes between the selection line and the control based on the interaction
between passage and treatment. To test the interaction, we fit two nested models
with and without the interaction term and compared them using an F-test with the
'anova' function. We then present the slopes for each treatment, which
represented the change in CH~4~ oxidation rate per passage as a response to
selection.

We estimated the proportion of variation in CH~4~ oxidation rate due to
variation in the microbiome as the regression of divergence between the positive
line and the control on the cumulative selection differential [@falconer1996].
Divergence was calculated as the mean CH~4~ oxidation rate of the positive
treatment minus the mean CH~4~ oxidation rate of the control in each
passage. The selection differential was calculated as the difference between
the mean of the three selected jars and the mean of all twelve jars in a passage.
Cumulative selection differential was calculated as the sum of the selection
differential from all preceding selection events. We then regressed cumulative
divergence on cumulative selection differential using the 'lm' function.

@@

We observed a response to artificial selection on whole-ecosystem soil CH~4~ oxidation
rate (`r figures("response", display = "cite")`; difference of slopes:
F~`r compare_response_mods[[2,4]]`,`r compare_response_mods[[2,2]]`~ = 
`r myround(compare_response_mods[[2,6]], 2)`,
p = `r myround(compare_response_mods[[2,7]], 2)`). At the start of the experiment,
the positive selection treatment had a mean CH~4~ oxidation rate that was
`r percent_less(initial = tidy(response_model)[[1, 2]], final = tidy(response_model)[[3, 2]] + tidy(response_model)[[1, 2]])`\%
lower than the control
`r print_lm(response_model, 3, 'difference of y-intercepts')`. 
There was no change in CH~4~ oxidation rate in the control over the
five passages `r print_lm(response_model, 2)`. By contrast, the selection
treatment had a `r myround(per_slope, 1)`\% increase in CH~4~ oxidation rate per
passage `r print_lm(response_model, 4)`.

@@

Median ASV richness decreased from `r myround(rare_median[1, 2], 1)` (`r myround(rare_median[1, 3], 1)`) in passage 2 to `r myround(rare_median[2, 2], 1)` (`r myround(rare_median[2, 3], 1)`) in passage 5 (Kruskal-Wallace test: $\chi^2$ = 
`r myround(kt[1,1], 1)`, df = `r kt[1,3]`, p < 0.001; pairwise Wilcoxon test: p < 0.001). However, there was no
difference in richness between the selection treatment and the control in
passage 2 or 5 (pairwise Wilcoxon test: p = `r myround(wt[1, 3], 1)`).  In addition, there was no
correlation between richness and CH~4~ oxidation rate across the two treatments in passage 5 (Spearman's rho = `r myround(s5[1,1], 1)`, p = `r myround(s5[1,3], 1)`).

@@

Bray-Curtis dissimilarity of the soil microbiome varied strongly by passage and
weakly by treatment with an interaction between passage and treatment (`r figures("beta", display = "cite")`). 
Passage explained `r myround(100*tidy_permanova[1, 4])`\% of the
variation in Bray-Curtis dissimilarity (F~`r tidy_permanova[1, 2]`,`r tidy_permanova[4, 2]`~ = `r myround(tidy_permanova[1, 5], 1)`, p = `r myround(tidy_permanova[1, 6], 3)`), treatment explained 
`r myround(100*tidy_permanova[2, 4])`\%
of the variation (F~`r tidy_permanova[2, 2]`,`r tidy_permanova[4, 2]`~ = `r myround(tidy_permanova[2, 5], 1)`, p = `r myround(tidy_permanova[2, 6], 3)`), and the interaction
between treatment and passage explained 
`r myround(100*tidy_permanova[3, 4])`\% of the variation (F~`r tidy_permanova[3, 2]`,`r tidy_permanova[4, 2]`~ = `r myround(tidy_permanova[3, 5], 1)`, p = `r myround(tidy_permanova[3, 6], 3)`). There was 
no difference in dispersion between treatments or passages (F~3,44~ = 0.91, p = 0.45). 
Finally, CH~4~ oxidation rate was correlated with Bray-Curtis dissimilarity across both treatments in passage 5 and explained `r myround(dbrda_percent_explained[1], 1)`\% of the variation in Bray-Curtis dissimilarity (dbRDA: F~`r dbrda_aov[1, 1]`,`r dbrda_aov[2, 1]`~ = `r myround(dbrda_aov[1, 3], 2)`, p = `r myround(dbrda_aov[1, 4], 3)`)

@@

To identify taxa that responded
to selection on soil CH~4~ oxidation rate, we tested the differential relative
abundance of families in the selected jars relative to the control jars within
passage 5 using three methods and then plotted the taxa identified by all three methods. We identified `r length(da_fams)`
families that were enriched or depleted in the selection treatment relative to
the control (`r figures("da", display = "cite")`).

Overall, none of the families enriched in the selection treatment
contain known methanotrophs. Several taxa identified had a 
higher taxonomic designation that contains methanotrophs, for example, the
Gammaproteobacteria class had a large effect size.
The Gammaproteobacteria include the type I and type X methanotrophs in the families *Methylococcaceae* and *Methylothermaceae* [@stein2012].
However, the Gammaproteobacteria is among the most diverse 
groups in the Prokaryotes, so this is not strong evidence for a selection response
by methanotrophs [@garrity2005].
In addition, the *Puniceicoccaceae* is a member of the phylum Verrucomicrobia. The Verrucomicrobia is a diverse group that contain
known methanotrophs as well as ammonia-oxidizing bacteria [@freitag2003]. Other
than these two groups, none of the other taxa enriched in the selection treatment
are known to be related to methanotrophs. Two groups in the Armatimonadales were
enriched in the selection treatment including the family *Fimbriimonadaceae* and
an unclassified ASV from the order Armatimonadales [@im2012].
*Cytophagaceae* was also enriched in
the selection treatment and contains a number of mainly aerobic heterotrophs
that can digest a variety of macromolecules [@mcbride2014]. 
The remaining families include the uncultured family 0319-6G20, *Diplorickettsiaceae*, *Rhodospirillaceae*, and an unclassified Kapabacteriales.

We did not identify any methanotrophic families in the overall differential abundance analysis. 
However, we wanted to look more closely at the known methanotrophs in our dataset to be sure that they did not have an effect.
To do this, we subset the ASVs in our dataset that were in families that contained methanotrophs.
Only two families were represented: *Methylacidiphilaceae* and *Beijerinckiaceae*. 
Aggregating reads at the family level, neither family was differentially abundant between the two treatments.
However, aggregated at the genus level, a group of unclassified genera in the *Beijerinckiaceae* were depleted in the selection treatment and the genus
*Rhodoblastus*, a member of the *Beijerinckiaceae*, was enriched in the
selection treatment.  While many *Beijerinckiaceae* are methanotrophs, several
taxa in this family have lost the ability to oxidize CH~4~ and it appears
that *Rhodoblastus* species are not able to grow on CH~4~, though they can grow
on methanol [@tamas2014; @dedysh2016]. Based on this analysis,
it appears that no methanotrophs were enriched in the positive selection treatment.
